---
title: "TMS Analysis & Results -- One Sided Tests"
author: "Aadi Kalloo"
date: "March 28, 2016"
output:
  html_document:
    toc: true
    toc_float: true
---
This document shows the complete process of analysis for TMS Data.

These are options that are most likely to need adjustment by users:
```{r}
#Data: https://github.com/aadikalloo/BridgeR/blob/master/tmsdata_MARCH28.csv
URL = "https://raw.githubusercontent.com/aadikalloo/BridgeR/master/tmsdata_MARCH28.csv"

#two-sided or one-sided tests for control vs rls
t_test_type   = "one.sided"  #options are: "two.sided", "one.sided"
mwu_test_type = "one.sided"  #options are: "two.sided", "one.sided"

#two-sided or one-sided tests for male vs female
t_test_typeMF   = "one.sided"  #options are: "two.sided", "one.sided"
mwu_test_typeMF = "one.sided"  #options are: "two.sided", "one.sided"

#specify number of digits to display in output tables
numdigits = 4

```

Defining various constants for code later on:
```{r}
t_test_start_col = 5
t_test_end_col = 26 #tms measures go until column 12; columns 13-26 contains RMT/intensity data

mwu_start_col = 5
mwu_end_col = 26 #tms measures go until column 12; columns 13-26 contains RMT/intensity data

data_start_col = 5
data_end_col = 26

ttestdata_end_col = t_test_end_col - t_test_start_col - 1
```

#Analysis


##Load Data

Load necessary packages, pull data from GitHub, and set up as dataframe:
```{r, warning = FALSE, message = FALSE}
#install.packages("stringr")
#install.packages("tidyr")
#install.packages("dplyr")
#install.packages("knitr")
#install.packages("pander")

library(stringr)
library(tidyr)
library(dplyr)
library(knitr)
library(pander)


tmsdata = read.csv(URL)
tmsdata2 = gather(tmsdata, "Measure", "Value", 5:26)
```

##Removal of participants

Define which subjects should be removed:
```{r, warning=FALSE}
#We remove subjects 3, 14, 21, 76 because they either had crazy outliers or missing data
subjects_to_remove = c("G13400", "G07396", "G76051", "G12737") 

#according to email on March 22 2016, we want to remove the following subjects for the following reasons:
# G80287 - Diabetes
# G17673 - Thyroid
# G29018 - Thyroid
# G23418 - Thyroid
# G39035 - Thyroid
# G75351 - Thyroid

subjects_to_remove = append(subjects_to_remove, c("G80287", "G17673", "G29018", "G23418", "G39035", "G75351"))

```

Remove subjects from dataframe:
```{r}
tmsdata = tmsdata[which(!tmsdata$ID %in% subjects_to_remove),]
```

##Removing Outliers

Comb through data values to remove outliers. Outliers based on 1.5*IQR:
```{r}
#Allow outliers to be removed broken down by variable and status
tmsdata_r = subset(tmsdata, tmsdata$STATUS == "RLS")
tmsdata_c = subset(tmsdata, tmsdata$STATUS == "Control")

for (var in data_start_col:data_end_col) {
  third_quartile = quantile(tmsdata_r[, var], na.rm = TRUE)[[4]]
  first_quartile = quantile(tmsdata_r[, var], na.rm = TRUE)[[2]]
  low_cut = first_quartile - 1.5 * IQR(tmsdata_r[, var], na.rm = TRUE)
  hi_cut  = third_quartile + 1.5 * IQR(tmsdata_r[, var], na.rm = TRUE)
  
  for (count in 1:length(tmsdata_r[, var])) {
    if(!is.na(tmsdata_r[count, var])) {
      if(tmsdata_r[count, var] < low_cut | tmsdata_r[count,var] > hi_cut) {
         tmsdata_r[count, var] = NA #Any value that is not within range is removed
      }
    }
  }
      
}



for (var in data_start_col:data_end_col) {
  third_quartile = quantile(tmsdata_c[, var], na.rm = TRUE)[[4]]
  first_quartile = quantile(tmsdata_c[, var], na.rm = TRUE)[[2]]
  low_cut = first_quartile - 1.5 * IQR(tmsdata_c[, var], na.rm = TRUE)
  hi_cut  = third_quartile + 1.5 * IQR(tmsdata_c[, var], na.rm = TRUE)
  
  for (count in 1:length(tmsdata_c[, var])) {
   if(!is.na(tmsdata_c[count, var])) {
    if(tmsdata_c[count, var] < low_cut | tmsdata_c[count,var] > hi_cut) {
      tmsdata_c[count, var] = NA #Any value that is not within range is removed
    }
   }
  }
      
}

#put data back together into a single dataframe
tmsdata = rbind(tmsdata_r, tmsdata_c)
```

##Functions for T tests

Create function that performs Independent Samples T Tests and organizes the data into a table:
```{r, warning=FALSE}

tmsTtest <- function(tmsdata1) {
  
  #initialize dataframe for t test data
  ttestdata1 = data.frame(1,1,1,1,1,1,1,1)
  
  #set column names:
  names(ttestdata1) = c("Measure",
                       "Control_count",
                       "RLS_count",
                       "Control_mean", 
                       "RLS_mean", 
                       "Control_median", 
                       "RLS_median", 
                       "pValue")
  
  #Independent T tests for tms data, ignores gender:
  for (var in t_test_start_col:t_test_end_col) {
    mean_c = mean(subset(tmsdata1[, var],   tmsdata1$STATUS == "Control"), na.rm=TRUE)
    mean_r = mean(subset(tmsdata1[, var],   tmsdata1$STATUS == "RLS"), na.rm=TRUE)
    
    if (t_test_type == "one.sided") {
      if (mean_c > mean_r) {t_test_type1 = "greater"}
      if (mean_c < mean_r) {t_test_type1 = "less"}
    } else {
      t_test_type1 = "two.sided"
    }
    
    t_test = #the t-test is performed on each variable
        t.test(subset(tmsdata1[, var], tmsdata1$STATUS == "Control"), 
               subset(tmsdata1[, var], tmsdata1$STATUS == "RLS"), 
               paired = FALSE, 
               alternative = t_test_type1,
               na.rm = TRUE) #%>% print()
  
    #set up the t-test-data table/dataframe:
    ttestdata1[var - t_test_start_col + 1,] = 
        c(    
          tmsdata1[var] %>% names(),
          length(subset(tmsdata1[, var], tmsdata1$STATUS == "Control" & tmsdata1[, var] != "NA")),
          length(subset(tmsdata1[, var], tmsdata1$STATUS == "RLS" & tmsdata1[, var] != "NA")),
          mean(subset(tmsdata1[, var],   tmsdata1$STATUS == "Control"), na.rm=TRUE),
          mean(subset(tmsdata1[, var],   tmsdata1$STATUS == "RLS"), na.rm=TRUE),
          median(subset(tmsdata1[, var], tmsdata1$STATUS == "Control"), na.rm=TRUE),
          median(subset(tmsdata1[, var], tmsdata1$STATUS == "RLS"), na.rm=TRUE),
          t_test$p.value
        )
    
  }
  return(ttestdata1)
}


#This function compares male and female data for a given RLS Status
tmsTtest2 <- function(tmsdata1) {
  
  #initialize dataframe for t test data
  ttestdata1 = data.frame(1,1,1,1,1,1,1,1)
  
  #set column names:
  names(ttestdata1) = c("Measure",
                       "Male_count",
                       "Female_count",
                       "Male_mean", 
                       "Female_mean", 
                       "Male_median", 
                       "Female_median", 
                       "pValue")
  
  #Independent T tests for tms data, separating by gender:
  for (var in t_test_start_col:t_test_end_col) {
  
    mean_c1 = mean(subset(tmsdata1[, var],   tmsdata1$GENDER == "Male"),   na.rm=TRUE)
    mean_r1 = mean(subset(tmsdata1[, var],   tmsdata1$GENDER == "Female"), na.rm=TRUE)
    
    if (t_test_typeMF == "one.sided") {
      if (mean_c1 > mean_r1) {t_test_typeMF1 = "greater"}
      if (mean_c1 < mean_r1) {t_test_typeMF1 = "less"}
    } else {
      t_test_typeMF1 = "two.sided"
    }
    
    
    t_test = #the t-test is performed on each variable
        t.test(subset(tmsdata1[, var], tmsdata1$GENDER == "Male"), 
               subset(tmsdata1[, var], tmsdata1$GENDER == "Female"), 
               paired = FALSE, 
               alternative = t_test_typeMF1,
               na.rm = TRUE) #%>% print()
  
    #set up the t-test-data table/dataframe:
    ttestdata1[var - t_test_start_col + 1,] = 
        c(    
          tmsdata1[var] %>% names(),
          length(subset(tmsdata1[, var], tmsdata1$GENDER == "Male" & tmsdata1[, var] != "NA")),
          length(subset(tmsdata1[, var], tmsdata1$GENDER == "Female" & tmsdata1[, var] != "NA")),
          mean(subset(tmsdata1[, var],   tmsdata1$GENDER == "Male"), na.rm=TRUE),
          mean(subset(tmsdata1[, var],   tmsdata1$GENDER == "Female"), na.rm=TRUE),
          median(subset(tmsdata1[, var], tmsdata1$GENDER == "Male"), na.rm=TRUE),
          median(subset(tmsdata1[, var], tmsdata1$GENDER == "Female"), na.rm=TRUE),
          t_test$p.value
        )
    
  }
  return(ttestdata1)
}
```

##Functions for Mann Whitney U Tests

Create function that performs Mann Whitney U Tests and organizes the data into a table:

```{r, warning = FALSE}

tmsMWUtest <- function(tmsdata) {
  
  #initialize dataframe for mwu data :
  mwutestdata = data.frame(1,1,1,1,1,1,1,1)
  
  #set column names:
  names(mwutestdata) = c("Measure",
                         "Control_count",
                         "RLS_count",
                         "Control_mean", 
                         "RLS_mean", 
                         "Control_median", 
                         "RLS_median", 
                         "pValue")
  
  #Mann-Whitney U tests for tms data, ignores gender:
  for (var in mwu_start_col:mwu_end_col) {
    median_c = median(subset(tmsdata[, var],   tmsdata$STATUS == "Control"), na.rm=TRUE)
    median_r = median(subset(tmsdata[, var],   tmsdata$STATUS == "RLS"), na.rm=TRUE)
    
    if (mwu_test_type == "one.sided") {
      if (median_c > median_r) {mwu_test_type1 = "greater"}
      if (median_c < median_r) {mwu_test_type1 = "less"}
    } else {
      mwu_test_type1 = "two.sided"
    }
    
    
    
    mwu_test = 
    wilcox.test(subset(tmsdata[, var], tmsdata$STATUS == "Control"), 
                subset(tmsdata[, var], tmsdata$STATUS == "RLS"), 
                paired = FALSE, 
                alternative = mwu_test_type1,
                correct = FALSE,
                na.rm = TRUE) #%>% print()
    
    
    #set up table/dataframe for mwu test data:
    mwutestdata[var - t_test_start_col + 1,] = 
      c(    
        tmsdata[var] %>% names(),
        length(subset(tmsdata[, var], tmsdata$STATUS == "Control" & tmsdata[, var] != "NA")),
        length(subset(tmsdata[, var], tmsdata$STATUS == "RLS" & tmsdata[, var] != "NA")),
        mean(subset(tmsdata[, var],   tmsdata$STATUS == "Control"), na.rm=TRUE),
        mean(subset(tmsdata[, var],   tmsdata$STATUS == "RLS"), na.rm=TRUE),
        median(subset(tmsdata[, var], tmsdata$STATUS == "Control"), na.rm=TRUE),
        median(subset(tmsdata[, var], tmsdata$STATUS == "RLS"), na.rm=TRUE),
        mwu_test$p.value
      )
  }
  return(mwutestdata)
}

#This function compares male and female data for a given RLS Status
tmsMWUtest2 <- function(tmsdata) {
  
  #initialize dataframe for mwu data :
  mwutestdata = data.frame(1,1,1,1,1,1,1,1)
  
  #set column names:
  names(mwutestdata) = c("Measure",
                         "Male_count",
                         "Female_count",
                         "Male_mean", 
                         "Female_mean", 
                         "Male_median", 
                         "RLS_median", 
                         "pValue")
  
  #Mann-Whitney U tests for tms data, separating by Gender:
  for (var in mwu_start_col:mwu_end_col) {
    median_c = median(subset(tmsdata[, var],   tmsdata$GENDER == "Male"), na.rm=TRUE)
    median_r = median(subset(tmsdata[, var],   tmsdata$GENDER == "Female"), na.rm=TRUE)
    
    if (mwu_test_typeMF == "one.sided") {
      if (median_c > median_r) {mwu_test_typeMF1 = "greater"}
      if (median_c < median_r) {mwu_test_typeMF1 = "less"}
    } else {
      mwu_test_typeMF1 = "two.sided"
    }
    
    
    
    mwu_test = 
    wilcox.test(subset(tmsdata[, var], tmsdata$GENDER == "Male"), 
                subset(tmsdata[, var], tmsdata$GENDER == "Female"), 
                paired = FALSE, 
                alternative = mwu_test_typeMF1,
                correct = FALSE,
                na.rm = TRUE) #%>% print()
    
    
    #set up table/dataframe for mwu test data:
    mwutestdata[var - t_test_start_col + 1,] = 
      c(    
        tmsdata[var] %>% names(),
        length(subset(tmsdata[, var], tmsdata$GENDER == "Male" & tmsdata[, var] != "NA")),
        length(subset(tmsdata[, var], tmsdata$GENDER == "Female" & tmsdata[, var] != "NA")),
        mean(subset(tmsdata[, var],   tmsdata$GENDER == "Male"), na.rm=TRUE),
        mean(subset(tmsdata[, var],   tmsdata$GENDER == "Female"), na.rm=TRUE),
        median(subset(tmsdata[, var], tmsdata$GENDER == "Male"), na.rm=TRUE),
        median(subset(tmsdata[, var], tmsdata$GENDER == "Female"), na.rm=TRUE),
        mwu_test$p.value
      )
  }
  return(mwutestdata)
}
```

##Preparing Data

Call all functions:
```{r, warning = FALSE}
ttestdata   = tmsTtest(tmsdata)
mwutestdata = tmsMWUtest(tmsdata)

ttestdataMF_Control   = tmsTtest2(subset(tmsdata, tmsdata$STATUS=="Control"))
ttestdataMF_RLS       = tmsTtest2(subset(tmsdata, tmsdata$STATUS=="RLS"))

mwutestdataMF_Control = tmsMWUtest2(subset(tmsdata, tmsdata$STATUS=="Control"))
mwutestdataMF_RLS     = tmsMWUtest2(subset(tmsdata, tmsdata$STATUS=="RLS"))

ttestdata_males   = tmsTtest(subset(tmsdata, tmsdata$GENDER=="Male"))
mwutestdata_males = tmsMWUtest(subset(tmsdata, tmsdata$GENDER=="Male"))

ttestdata_females   = tmsTtest(subset(tmsdata, tmsdata$GENDER=="Female"))
mwutestdata_females = tmsMWUtest(subset(tmsdata, tmsdata$GENDER=="Female"))

```

```{r}

pValTable = data.frame(
  ttestdata$Measure,
  ttestdata$pValue,
  mwutestdata$pValue,
  ttestdata_males$pValue,   
  mwutestdata_males$pValue, 
  ttestdata_females$pValue,   
  mwutestdata_females$pValue, 
  ttestdataMF_Control$pValue,
  mwutestdataMF_Control$pValue, 
  ttestdataMF_RLS$pValue,    
  mwutestdataMF_RLS$pValue,
  stringsAsFactors = FALSE
)

names(pValTable) = c("Measure",
                     "All Control vs RLS,  TT",
                     "All Control vs RLS, MWU",
                     "Male Control vs RLS,  TT",
                     "Male Control vs RLS, MWU",                     
                     "Female Control vs RLS,  TT",
                     "Female Control vs RLS, MWU",
                     "Control Males vs Females,  TT",
                     "Control Males vs Females, MWU",
                     "RLS Males vs Females,  TT",
                     "RLS Males vs Females, MWU"                     
                    )


```



Convert values from chr to numeric:
```{r}

ttestdata[,2:length(names(ttestdata))] = sapply(ttestdata[,2:length(names(ttestdata))], as.numeric)
mwutestdata[,2:length(names(mwutestdata))] = sapply(mwutestdata[,2:length(names(mwutestdata))], as.numeric)

ttestdataMF_Control[,2:length(names(ttestdataMF_Control))] = sapply(ttestdataMF_Control[,2:length(names(ttestdataMF_Control))], as.numeric)
mwutestdataMF_Control[,2:length(names(mwutestdataMF_Control))] = sapply(mwutestdataMF_Control[,2:length(names(mwutestdataMF_Control))], as.numeric)

ttestdataMF_RLS[,2:length(names(ttestdataMF_RLS))] = sapply(ttestdataMF_RLS[,2:length(names(ttestdataMF_RLS))], as.numeric)
mwutestdataMF_RLS[,2:length(names(mwutestdataMF_RLS))] = sapply(mwutestdataMF_RLS[,2:length(names(mwutestdataMF_RLS))], as.numeric)

ttestdata_males[,2:length(names(ttestdata_males))] = sapply(ttestdata_males[,2:length(names(ttestdata_males))], as.numeric)
mwutestdata_males[,2:length(names(mwutestdata_males))] = sapply(mwutestdata_males[,2:length(names(mwutestdata_males))], as.numeric)

ttestdata_females[,2:length(names(ttestdata_females))] = sapply(ttestdata_females[,2:length(names(ttestdata_females))], as.numeric)
mwutestdata_females[,2:length(names(mwutestdata_females))] = sapply(mwutestdata_females[,2:length(names(mwutestdata_females))], as.numeric)

pValTable[,2:length(names(pValTable))] = sapply(pValTable[,2:length(names(pValTable))], as.numeric)

```

#Tables

##T-tests, All Control vs RLS
```{r}
emphasize.strong.cells(which(ttestdata < 0.05, arr.ind = TRUE))   #highlight values <0.05
pander(ttestdata, digits = numdigits, split.table = Inf)

```

##Mann Whitney U, All Control vs RLS
```{r}
emphasize.strong.cells(which(mwutestdata < 0.05, arr.ind = TRUE)) #highlight values <0.05
pander(mwutestdata, digits = numdigits, split.table = Inf)
```

##T-tests, Male Control vs RLS
```{r}
emphasize.strong.cells(which(ttestdata_males < 0.05, arr.ind = TRUE))   #highlight values <0.05
pander(ttestdata_males, digits = numdigits, split.table = Inf)

```

##Mann Whitney U, Male Control vs RLS
```{r}
emphasize.strong.cells(which(mwutestdata_males < 0.05, arr.ind = TRUE)) #highlight values <0.05
pander(mwutestdata_males, digits = numdigits, split.table = Inf)
```

##T-tests, Female Control vs RLS
```{r}
emphasize.strong.cells(which(ttestdata_females < 0.05, arr.ind = TRUE))   #highlight values <0.05
pander(ttestdata_females, digits = numdigits, split.table = Inf)

```

##Mann Whitney U, Female Control vs RLS
```{r}
emphasize.strong.cells(which(mwutestdata_females < 0.05, arr.ind = TRUE)) #highlight values <0.05
pander(mwutestdata_females, digits = numdigits, split.table = Inf)
```

##T-tests, Control Male vs Female
```{r}
emphasize.strong.cells(which(ttestdataMF_Control < 0.05, arr.ind = TRUE))   #highlight values <0.05
pander(ttestdataMF_Control, digits = numdigits, split.table = Inf)

```

##Mann Whitney U, Control Male vs Female
```{r}
emphasize.strong.cells(which(mwutestdataMF_Control < 0.05, arr.ind = TRUE)) #highlight values <0.05
pander(mwutestdataMF_Control, digits = numdigits, split.table = Inf)
```

##T-tests, RLS Male vs Female
```{r}
emphasize.strong.cells(which(ttestdataMF_RLS < 0.05, arr.ind = TRUE))   #highlight values <0.05
pander(ttestdataMF_RLS, digits = numdigits, split.table = Inf)

```

##Mann Whitney U, RLS Male vs Female
```{r}
emphasize.strong.cells(which(mwutestdataMF_RLS < 0.05, arr.ind = TRUE)) #highlight values <0.05
pander(mwutestdataMF_RLS, digits = numdigits, split.table = Inf)

```

##Combined Table of P-Values
```{r}
emphasize.strong.cells(which(pValTable < 0.05, arr.ind = TRUE)) #highlight values <0.05
pander(pValTable, digits = numdigits, split.table = Inf)

```


